Informing policy via dynamic models: Cholera in Haiti

### 2024-01-15 r1/response.tex

While the arguments provided throughout the article are well-articulated, there needs to be more supporting literature at the beginning of major sections. For instance, I don’t need to be convinced that the structure of a model should be based on a realistic theory about the observed phenomenon; namely, models should be a white box. However, not everyone is on board with this premise, and supporting literature that argues in favour of this approach should be mentioned. In short, more citations should be added at the beginning of each major section.
}

\report{
\textbf{\underline{2. Limitations}}

In various passages of this paper, it is hinted that modellers should refrain from deterministic models and instead opt for more realistic stochastic representations. While the critique of ODE models is valid, the shift to stochastic structures is not a free choice. For example, introducing extra-demographic variability adds one additional parameter (infinitesimal variance). In ODE models, one extra parameter can lead to unidentifiability, and there's no apparent reason why this would differ in a stochastic version. Moreover, transitioning to stochastic models involves abandoning well-established MCMC algorithms in favour of methods still in development. Therefore, modellers should not assume that more realistic models with additional parameters are necessarily better without proper caution. In my experience, diagnosing unidentifiability is easier in ODE models than in POMP structures. Hence, there is a trade-off between benefits and costs.
}

The referee raises several points in this paragraph, and we consider them one at a time.

We agree that adding to model complexity has costs, including loss of identifability, AIC and other formal statistical methods can assess whether the additional complexity is justified by a sufficiently large improvement of statistical fit. Identifiability and model simplicity have practical scientific value which goes beyond their statistical role in reducing generalization error and hence improving out-of-fit predictions. This is a variation on the issue of "statistical significance versus practical significance" familiar from introductory statistics textbooks. 
HOW DOES THIS PLAY OUT IN THE MS? HOW DO WE CLARIFY IT?

Moving from an ODE model to a stochastic model may not involve any additional parameters, if only demographic stochasticity is included.
Generally, it is advisable to consider the possibility of over-dispersed variation on the transmission rate \cite{breto09,he10} leading to one extra parameter compared to a deterministic model.
In the context of cholera dynamics, \cite{lemaitre19} found a substantial improvement in model fit by including stochasticity, involving one over-dispersion parameter.
stocks20 found similar results, favoring a stochastic dynamic model with over-dispersion, when investigating transmission dynamics of a different enteric pathogen, rotavirus.
In other contexts, king15 noted various scientific advantages for including stochasticity in the model; it is undoubtedly part of the biological dynamics, and its omission can lead to over-confident predictions.
In our study, we find that the deterministic model (model 2) has poor statistical fit.
This can lead to a risk of identifying spurious scientific phenomena: addition deterministic complexity added to the model may help to describe unmodeled stochastic phenomena present in the data, and may therefore spuriously be assessed as a highly statistically significant model improvement.

The decision on whether to choose Markov chain Monte Carlo (MCMC) methods is not fully aligned with the decision on whether to include stochasticity in the model.
In our literature review, of the 9 previous Haiti cholera investigations including stochastic dynamic models \cite{tuite11,mukandavire13,lewnard16,kunkel17,sallah17,azman12,kuhn14,azman15,lee20}, four of them carried out inference via MCMC \cite{lewnard16,sallah17,azman12,kuhn14}.
WHAT SORTS OF MCMC ALGORITHMS DID THEY USE?
We agree with the referee that MCMC is more routine for determinstic models than for stochastic models, and advanced MCMC methods such as Hamiltonian Monte Carlo become more readily applicable. 
Sequential Monte Carlo (SMC) was used for only two of the investigations in our literature review \cite{lee20,azman15}.
In principle, MCMC methods can be applied for inference in over-dispersed stochastic dynamic models via so-called particle MCMC (PMCMC) \cite{andrieu10}.
In practice, these models have been fitted by maximum likelihood, some examples including \cite{lemaitre19,fox22,molodecky23,pons-salort18,subramanian20,stocks20}.
In this case, identifiability is generally addressed via profile likelihood methods.

We have included some additional sentences to address these comments in the discussion section of the revision.
TODO

We acknowledge the existence of useful mathematical tools for studying ODE models, their identifiability and stability and limiting properties.
Stochastic dynamic models can be reduced to a deterministic model (a so-called determinsitic skeleton) which can usefully be analyzed to help understand the stochastic dynamics \cite{coulson04}.
We therefore encourage the use of ODE analysis tools whether or not a stochastic model is adopted.
We have not followed that approach in the current manuscript.

In summary, maximum likelihood (ML) inference via SMC nowaways provides a fairly well estabilished alternative to Bayesian MCMC inference. ML appears to be methodologically more convenient for stochastic dynamic models with overdispersion. When analyzing long epidemiological datasets, this stochasticity appears to be important for statistical model fit. That alone is not enough to show that models with overdispersed stochastic dynamics are scientifically preferable. However, discovery of a  model feature which generates a large improvement in statistcal fit from adding one more parameter is typically treated as scientific evidence favoring the inclusion of that feature in a scientific model for the phenomenon.

### 2024-01-15 ms.Rnw
response to R2 about mcmc, etc

%%%%%% EDITING MACROS %%%%%%%%%
% orange for EI
\definecolor{orange}{rgb}{1,0.5,0}
\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
\newcommand\eic[1]{\textcolor{orange}{[#1]}}
% green for JW
\definecolor{green}{rgb}{0,0.5,0}
\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
\newcommand\jwc[1]{\textcolor{green}{[#1]}}
% purple for JJ
\definecolor{purple}{rgb}{0.5,0,1}
\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
% cyan for AR
\definecolor{cyan}{rgb}{0,.5,.5}
\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
\newcommand\arc[1]{\textcolor{cyan}{#1}}
% light brown for KT
\definecolor{lightbrown}{rgb}{0.5,0.5,0}
\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}

@@

\section*{Abstract}
Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
Thus, we encounter the methodological task of building credible, data-informed decisions based on stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
As a case study, we consider the 2010-2019 cholera epidemic in Haiti.
We study three dynamic models developed by expert teams to advise on vaccination policies.
We assess previous methods used for fitting and evaluating these models, and we develop data analysis strategies leading to improved statistical fit.
Specifically, we present approaches to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.
Our workflow is reproducible and extendable, facilitating future investigations of this disease system.

>>>>

@@

\section*{Abstract}
Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
Thus, we encounter the methodological task of building credible, data-informed decisions based on stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
\ei{I'VE RE-WRITTEN HERE, LOOKING TO BE MORE SPECIFIC ABOUT THE GOALS AND CONTRIBUTION OF THE PAPER.} 
We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti.
We consider three dynamic models developed by expert teams to advise on vaccination policies.
We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.
Specifically, we present approaches to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.
Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
